# Exercise 3


```{r}
date()
```

In  this exercise, I study data obtained from seconday education students in two Portuguese schools. The data attributes include student grades, demographic, social and school related features) and it was collected by using school reports and questionnaires. Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). I have joined the two datasets into one joined set and will work with that.

Here is the dataset I use:
```{r}
alc_student<- read.csv("data/studentclean.csv", header = TRUE, sep = ",")
names(alc_student)
```
**Explanations for the variables:**
1 school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)
2 sex - student's sex (binary: 'F' - female or 'M' - male)
3 age - student's age (numeric: from 15 to 22)
4 address - student's home address type (binary: 'U' - urban or 'R' - rural)
5 famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)
6 Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart)
7 Medu - mother's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
8 Fedu - father's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
9 Mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
10 Fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
11 reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')
12 guardian - student's guardian (nominal: 'mother', 'father' or 'other')
13 traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
16 schoolsup - extra educational support (binary: yes or no)
17 famsup - family educational support (binary: yes or no)
19 activities - extra-curricular activities (binary: yes or no)
20 nursery - attended nursery school (binary: yes or no)
21 higher - wants to take higher education (binary: yes or no)
22 internet - Internet access at home (binary: yes or no)
23 romantic - with a romantic relationship (binary: yes or no)
24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25 freetime - free time after school (numeric: from 1 - very low to 5 - very high)
26 goout - going out with friends (numeric: from 1 - very low to 5 - very high)
27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
29 health - current health status (numeric: from 1 - very bad to 5 - very good)
alc_use - average from Dalc and Walc
High_use TRUE if average alc_use is above 2


Next, I will choose 4 interesting variables that could have aconnection with the student alcohol consumption. 
I assume that if the student is not in a romantic relationship (romantic = no), they use more alcohol, as they go out more often and are more eager to meet new people. 
Further, I assume that going out of with friends (goout) as a highly positive correlation with alcohol consumption.
I hypothesize that fathers education (Fedu) has a negative correlation with the alcohol usage, although this is bit of a strecth. Interesting to see the results. 
Weekly studytime (studytime) definitely has a negative correlation with alcohol usage. 

**Let us first study graphically whether my hypotheses are correct:**

```{r}

# Summary of alc_use
summary(alc_student$alc_use)
summary(alc_student$high_use)

# Cross-tabulation for each categorical variable with alc_use
table(alc_student$romantic, alc_student$alc_use)
table(alc_student$goout, alc_student$alc_use)
table(alc_student$Fedu, alc_student$alc_use)
table(alc_student$studytime, alc_student$alc_use)


```
Out of 370 observations, we have 111 high alcohol users. The median for alcohol consumption is 1.5/day, which imply that the vast majority is in low alcohol consumption group.

**Let us try bar plots:**
```{r}
library(ggplot2)

# Bar plot for each categorical variable against alc_use
ggplot(alc_student, aes(x = romantic)) + 
  geom_bar() + 
  facet_wrap(~high_use)

ggplot(alc_student, aes(x = goout)) + 
  geom_bar() + 
  facet_wrap(~high_use)

ggplot(alc_student, aes(x = Fedu)) + 
  geom_bar() + 
  facet_wrap(~high_use)

ggplot(alc_student, aes(x = studytime)) + 
  geom_bar() + 
  facet_wrap(~high_use)


```

Based on the graphical representation of the data relationships, it seems that my hypotheses does not apply the way I hoped. Only the frequency of going out with friends seems to have a positive relationship with the high alcohol consumption.

**Let us next continue the analysis by logistic regression using a generalized linear model:**

```{r}
# Logistic regression model
logistic_model <- glm(high_use ~ romantic + goout + Fedu + studytime, 
                      data = alc_student, 
                      family = "binomial")
# Model summary
summary(logistic_model)

```
I use the generalized linear model (GLM), which should be more suitable when the dependent variable is binary. It seems like studytime and goout are statistically significant, where goout has a positive correlation with high alcohol use and studytime has a negative correlation. Both effects are in line with my original hypotheses. Relationship and fathers education does not have any statistical significance.

Next, we **present and interpret the coefficients of the model as odds ratios and provide confidence intervals for them.**

```{r}
# Exponentiate the coefficients to get odds ratios
odds_ratios <- exp(coef(logistic_model))
print(odds_ratios)

# Calculate and exponentiate confidence intervals
conf_int <- exp(confint(logistic_model))
print(conf_int)

```
The odds ratio greater than 1 indicates that the predictor increases the odds of being an high alcohol user. The results show that the 'goout' variable increases the odds of TRUE answer in high_use, while 'studytime' decreases the odds of TRUE answer. Similarly, 'Fedu' would increase and 'romantic ->yes' would decrease the odds of being a high alcohol user, but their condifence intervals range below and above one, implying that the effect is not clear.
For 'goout' and 'studytime', the confidence intervals imply that there is a 95% change that the variables increase/decrease the odds of high alcohol use. 

Next, we continue the analysis using 'goout' and 'studytime' and explore the predictive power of our model.

```{r}

# Predict probabilities
predicted_probs <- predict(logistic_model, type = "response")

# Classify as TRUE or FALSE based on a 0.5 cutoff
predicted_class <- ifelse(predicted_probs > 0.5, TRUE, FALSE)

alc_student$predicted_high_use <- predicted_class

# 2x2 Cross Tabulation
table(Actual = alc_student$high_use, Predicted = alc_student$predicted_high_use)


# Training error
training_error <- mean(alc_student$high_use != alc_student$predicted_high_use)
print(training_error)

library(ggplot2)

ggplot(alc_student, aes(x = high_use, fill = as.factor(predicted_high_use))) +
  geom_bar(position = "dodge") +
  labs(x = "Actual High Use", fill = "Predicted")



```

So, from 259 FALSE answers, my model was able to predict 238 correctly, while it incorrectly predicted 21 of them as TRUE. From 111 TRUE answers, it got 41 correct and predicted 70 as false. 

Let us compare our results with a simple guessing strategy:

```{r}
# Simple guessing strategy: classify all as the most frequent class
most_frequent_class <- ifelse(mean(alc_student$high_use) > 0.5, TRUE, FALSE)
simple_guess_accuracy <- mean(alc_student$high_use == most_frequent_class)

# Compare
model_accuracy <- 1 - training_error
print(c("Model Accuracy" = model_accuracy, "Simple Guess Accuracy" = simple_guess_accuracy))


```

The accuracy of my model is ~0.754 and a simple guessing strategy gives 0.7 accuracy. This is not a very convincing result, and I must admit that my model is not very accurate in predicting the outcome.

**BONUS:**
Let us next perform a 10-fold cross-validation on our model. 

```{r}
library(caret)

set.seed(123) # for reproducibility
train_control <- trainControl(method = "cv", number = 10)

# Convert 'high_use' from logical to factor
alc_student$high_use <- as.factor(alc_student$high_use)

# Now, run the 10-fold cross-validation
model_cv <- train(high_use ~ goout + studytime,
                  data = alc_student,
                  method = "glm",
                  family = "binomial",
                  trControl = train_control)

# Look at the results
print(model_cv)


```
The Kappa statistic (or Cohen's Kappa) is a measure of the agreement between the predicted and actual classifications, corrected for the agreement that could occur by chance.  My results 0.328 imply that the model result is better at predicting the 'high_use' status than a random chance. 








